---
title: "Exploratory Data Analysis Script"
author: "Ted Zerbe"
date: "December 19, 2022"
output:
  html_document:
    df_print: paged
---

## Front Matter

```{r, message=FALSE}
# clear environment
rm(list = ls())

# load packages
library(tidyverse)
library(ggcorrplot)
```

```{r}
# read in data
training <- read.csv("training.csv")
```

## Data Preparation

```{r}
# inspect data
glimpse(training)
```

```{r}
# inspect data
head(training)
```

```{r}
# check for missing values
colSums(is.na(training))

# check for NULL
colSums(training == 'NULL')
```

SpinRate is a character variable and must be converted to a numeric variable.

```{r, warning=FALSE}
# convert SpinRate to numeric variable
training <- training %>%
            mutate(SpinRate = as.numeric(SpinRate))
```

```{r}
# check for missing values
colSums(is.na(training))
```

```{r}
which(is.na(training$SpinRate))
```

```{r}
# convert InPlay to factor
training <- training %>%
            mutate(InPlay = as.factor(InPlay))
```

```{r}
# re-inspect data
glimpse(training)
```

I will now explore this dataset with descriptive statistics and appropriate plots. 

## Univariate Analysis

### Quantitative

```{r}
# plot histogram
ggplot(data = training, mapping = aes(x = Velo)) +
  geom_histogram(color = 'black', fill = 'steelblue1') +
  labs(x = 'Fastball Velocity')
```

```{r}
# calculate mean and sd
training %>%
  summarise(meanVelo = mean(Velo),
            sdVelo = sd(Velo))
```

```{r}
# find min, 25th percentile, median, 75th percentile, and max
fivenum(training$Velo)
```

```{r}
# detect outliers

# calculate IQR
IQR_velo = 95.66 - 92.54

# upper bound
95.66 + (1.5 * IQR_velo)

# lower bound
92.54 - (1.5 * IQR_velo)
```

```{r}
# visualize outliers
ggplot(data = training, mapping = aes(y = Velo)) +
  geom_boxplot()
```

Velocity is roughly normally distributed with a mean around 94 and a standard deviation of about 2.7. Values range between 102.4 and 59.76. There are a number of outliers to this distribution as well. The vast majority of them are on the lower tail.

```{r}
# plot histogram
ggplot(data = training, mapping = aes(x = SpinRate)) +
  geom_histogram(color = 'black', fill = 'steelblue1') +
  labs(x = 'Fastball Spin Rate')
```

```{r}
# calculate mean and sd
training %>%
  summarise(meanSpinRate = mean(SpinRate, na.rm = TRUE),
            sdSpinRate = sd(SpinRate, na.rm = TRUE))
```

```{r}
# find min, 25th percentile, median, 75th percentile, and max
fivenum(training$SpinRate)
```

```{r}
# detect outliers

# calculate IQR
IQR_spinrate = 2367 - 2107

# upper bound
2367 + (1.5 * IQR_spinrate)

# lower bound
2107 - (1.5 * IQR_spinrate)
```

```{r}
# visualize outliers
ggplot(data = training, mapping = aes(y = SpinRate)) +
  geom_boxplot()
```

SpinRate is roughly normally distributed with a mean around 2239 and a standard deviation of about 196. Values range between 3061 and 770. There are a number of outliers to this distribution as well.

```{r}
# plot histogram
ggplot(data = training, mapping = aes(x = HorzBreak)) +
  geom_histogram(color = 'black', fill = 'steelblue1') +
  labs(x = 'Fastball Horizontal Break')
```

```{r}
# calculate mean and sd
training %>%
  summarise(meanHorzBreak = mean(HorzBreak),
            sdHorzBreak = sd(HorzBreak))
```

```{r}
# find min, 25th percentile, median, 75th percentile, and max
fivenum(training$HorzBreak)
```

```{r}
# detect outliers

# calculate IQR
IQR_horzbreak = 13.60 - 5.73

# upper bound
13.60 + (1.5 * IQR_horzbreak)

# lower bound
5.73 - (1.5 * IQR_horzbreak)
```

```{r}
# visualize outliers
ggplot(data = training, mapping = aes(y = HorzBreak)) +
  geom_boxplot()
```

HorzBreak is roughly normally distributed with a mean around 9.5 and a standard deviation of about 5. Values range between 28.04 and -6.27. There are only two outliers in this distribution, one on each end.

```{r}
# plot histogram
ggplot(data = training, mapping = aes(x = InducedVertBreak)) +
  geom_histogram(color = 'black', fill = 'steelblue1') +
  labs(x = 'Fastball Vertical Break')
```

```{r}
# calculate mean and sd
training %>%
  summarise(meanVertBreak = mean(InducedVertBreak),
            medianVertBreak = median(InducedVertBreak),
            sdVertBreak = sd(InducedVertBreak))
```

```{r}
# find min, 25th percentile, median, 75th percentile, and max
fivenum(training$InducedVertBreak)
```

```{r}
# detect outliers

# calculate IQR
IQR_vertbreak = 17.63 - 11.36

# upper bound
17.63 + (1.5 * IQR_vertbreak)

# lower bound
11.36 - (1.5 * IQR_vertbreak)
```

```{r}
# visualize outliers
ggplot(data = training, mapping = aes(y = InducedVertBreak)) +
  geom_boxplot()
```

HorzBreak is roughly left-skewed with a median around 15 and a standard deviation of about 4.6. Values range between 24.86 and -6.82. There are a number of outliers to this distribution as well. All of the outliers are on the lower tail.

### Categorical

```{r}
# create bar chart
ggplot(data = training, mapping = aes(x = InPlay)) +
  geom_bar(color = 'black', fill = 'steelblue1') +
  labs(x = 'Fastball In Play?')
```

```{r}
# create frequency and relative frequency table
training %>%
  group_by(InPlay) %>%
  summarize(count = n(),
            prop = n()/nrow(training))
```

Over 2700, or about 27%, fastballs in this dataset were put in play.

## Bivariate Analysis

Now we will explore the relationships between the variables.

### Two Quantitative

```{r}
# make a dataset including only quantitative variables
quant <- training %>%
         select(-c(InPlay)) %>%
         drop_na()

# compute a correlation matrix
corr <- round(cor(quant), 3)
head(corr)
```

```{r}
ggcorrplot(corr, hc.order = TRUE, type = "lower",
     outline.col = "white")
```

The strongest linear correlation is between induced vertical break and horizontal break (-0.577).

```{r}
ggplot(data = training, mapping = aes(x = Velo, y = SpinRate)) +
  geom_point() +
  labs(x = 'Fastball Velocity', y = 'Fastball Spin Rate')
```

```{r}
ggplot(data = training, mapping = aes(x = HorzBreak, y = InducedVertBreak)) +
  geom_point() +
  labs(x = 'Fastball Horizontal Break', y = 'Fastball Vertical Break')
```

### One Quantitative, One Categorical

Now I will explore how the potential predictors relate to the target variable InPlay.

```{r}
ggplot(data = training, mapping = aes(x = InPlay, y = Velo)) +
  geom_boxplot()
```

```{r}
training %>%
  group_by(InPlay) %>%
  summarise(meanVelo = mean(Velo),
            sdVelo = sd(Velo))
```

```{r}
ggplot(data = training, mapping = aes(x = InPlay, y = SpinRate)) +
  geom_boxplot()
```

```{r}
training %>%
  group_by(InPlay) %>%
  summarise(meanSpinRate = mean(SpinRate, na.rm = TRUE),
            sdSpinRate = sd(SpinRate, na.rm = TRUE))
```

```{r}
ggplot(data = training, mapping = aes(x = InPlay, y = HorzBreak)) +
  geom_boxplot()
```

```{r}
training %>%
  group_by(InPlay) %>%
  summarise(meanHorzBreak = mean(HorzBreak),
            sdHorzBreak = sd(HorzBreak))
```

```{r}
ggplot(data = training, mapping = aes(x = InPlay, y = InducedVertBreak)) +
  geom_boxplot()
```

```{r}
training %>%
  group_by(InPlay) %>%
  summarise(meanVertBreak = mean(InducedVertBreak),
            sdVertBreak = sd(InducedVertBreak))
```

```{r}
training %>%
  mutate(InPlay = ifelse(InPlay == 1, 'Yes', 'No')) %>%
  group_by(InPlay) %>%
  summarise(meanVelo = mean(Velo),
            meanSpinRate = mean(SpinRate, na.rm = TRUE),
            meanHorzBreak = mean(HorzBreak),
            meanVertBreak = mean(InducedVertBreak))
```

